Demography and linked selection in wild and domesticated maize

### 2015-08-31 manuscript.tex
more edits

\subsection{Demography influences the efficiency of purifying selection}
\jri{section needs help} 
We find  a more pronounced decrease in nucleotide diversity around genes in teosinte compared to domesticated maize.
Because purifying selection is more efficient in larger populations  \cite{kimura1984},  this observation is consistent with our demographic model as well as previous work showing a smaller long-term effective population size in maize.  
This pattern of diversity is reversed, however, when we assess diversity at singleton polymorphisms.  
Under a simple neutral model, lower frequency polymorphisms will be younger and patterns of diversity of singleton polymorphisms should thus represent much more recent evolutionary history. 
We argue that the stronger decline in singleton diversity in and around genes in maize is the result of the increased efficacy of purifying selection  associated with the dramatic population expansion in maize following domestication.

This difference in recent $N_e$ should lead to an increased efficacy of purifying selection in the recent past of  Differences in diversity observed at  demographic estimation indicates dramatic the recent past, we expect for purifying selection to presently operate more efficiently in maize than in teosinte. This observation is very clearly demonstrated by the fact that singleton diversity in maize is more impacted by the distance to the nearest gene than it is in teosinte, as was shown in Figure \ref{fig:purify}.

A consequence of the inefficient purifying selection that maize experienced during its bottleneck is likely that it harbors more weakly deleterious alleles segregating at intermediate frequency than does teosinte. This could be a part of the explanation of why maize inbreds have continued to improve over the past several decades \cite{meghji1984}; if deleterious alleles tend to be recessive and are particularly frequent, they will have ample opportunities to display their phenotypes in inbreds. Our results also demonstrate that recent purifying selection in maize has become much more effective, potentially explaining the ongoing improvement of these inbreds as maize lines are continuously selected. Additionally, the large $N_e$ of modern maize compared to modern teosinte implies that for new mutations, selection will operate much more efficiently in maize.

 Importantly, our estimation of the parameters of the maize domestication bottleneck contribute to the understanding of how the demography of crop domestication can impact crop diversity. The bottleneck-effects from a sudden collapse in population size have been well studied and are known to impact crops for thousands of generations. Complementing this knowledge, our results demonstrate that the rapid expansion experienced by many crops after domestication can also have a profound influence on patterns of diversity, and the effects of this expansion should be accounted for as important contributors to long-term evolution. \jri{this is sorta fluff and doesn't say much}

\jri{need paragraph also talking about how this is important for considering impact of linked selection. cite some recent human papers that are considering both demography and linked selection, but also a number of plant papers that e.g. estimate the DFE from current data or just use the sny/nonsyn graphs and ignore recent demographic history. is capsella a good example? }

>>>>

\subsection{Demography influences the efficiency of purifying selection}
\jri{section needs help} 
We find  a more pronounced decrease in nucleotide diversity around genes in teosinte compared to domesticated maize (Figure \ref{fig:purify}A.
Because purifying selection is more efficient in larger populations  \cite{kimura1984},  this observation is consistent with our demographic model as well as previous work showing a smaller long-term effective population size in maize.  
Under a simple neutral model, lower frequency polymorphisms will be younger and patterns of diversity of singleton polymorphisms should  represent  more recent evolutionary history. 
In contrast to the pattern seen with pairwise diversity, we see a stronger decline of singleton diversity around genes in maize (Figure \ref{fig:purify}B.
We argue that the stronger decline in singleton diversity in and around genes in maize is the result of the increased efficacy of purifying selection  associated with the dramatic population expansion in maize following domestication.

A consequence of the inefficient purifying selection that maize experienced during its bottleneck is likely that it harbors more weakly deleterious alleles segregating at intermediate frequency than does teosinte. This could be a part of the explanation of why maize inbreds have continued to improve over the past several decades \cite{meghji1984}; if deleterious alleles tend to be recessive and are particularly frequent, they will have ample opportunities to display their phenotypes in inbreds. Our results also demonstrate that recent purifying selection in maize has become much more effective, potentially explaining the ongoing improvement of these inbreds as maize lines are continuously selected. Additionally, the large $N_e$ of modern maize compared to modern teosinte implies that for new mutations, selection will operate much more efficiently in maize. \jri{inbreds stuff OK, but need more. Chia, Rogers-Melnick, Mezmouk all looked at load. does this have implications for large effect load (i.e. should bneck have purged those?), for heterosis, architecture of quant. traits, etc.}

% Importantly, our estimation of the parameters of the maize domestication bottleneck contribute to the understanding of how the demography of crop domestication can impact crop diversity. The bottleneck-effects from a sudden collapse in population size have been well studied and are known to impact crops for thousands of generations. Complementing this knowledge, our results demonstrate that the rapid expansion experienced by many crops after domestication can also have a profound influence on patterns of diversity, and the effects of this expansion should be accounted for as important contributors to long-term evolution. \jri{this is sorta fluff and doesn't say much}

\jri{these results have implications for lots of studies. recent demographic history often ignored (examples). in crops, bneck is considered and loss of diversity, but expansion also likely ubiquitous and consequences not well studied.  in humans XYZ, and we see some of that already here. this argues that a full understanding of diversity in both wild and cultivated plants will require recent demography.}
